---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(forcats) # Library used to merge levels of factor variable.
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `brfss2013`. Delete this note when before you submit 
your work. 

```{r load-data}
load("brfss2013.RData")
```

### Introduction

<!-- After your introduction (which can be written later) data section should be fairly straightforward. Something along the lines of “I investigated these questions using the BRFSS dataset. This was collected from POPULATION using METHOD. Results ARE / ARE NOT generalizable to ANOTHER?POPULATION. STATEMENT ABOUT CAUSALITY" is sufficient. (I’m pretty confident you know the answers to fill in those blanks.) You will also need to address potential biases. They don't have to be actually present in the data, but given BRFSS codebook's information you should be able to find 2 or 3 things to keep in mind as you use their data. Do not just copy the BRFSS codebook without rewriting it in your own words and explaining what it means.  -->

<!-- 2 pt for correct reasoning for generabizability – Answer should discuss whether random sampling was used. Learners might discuss any reservations, those should be well justified. -->

<!-- 1 pt for correct reasoning for causality – Answer should discuss whether random assignment was used. -->

* * *

## Part 1: Data
The BRFSS dataset is an observational study of data collected from the United States non-institutionalized adult population, aged 18 years or older, all over it's 50 states. Data is collected in a state-specific basis, both prospectively (as events unfold) and retrospectively (after events have taken place, i.e. monthly data collection). At first they gathered data through surveys made by landline telephone calls; using random digit dialing, asking questions to a randomnly selected adult in the house-hold (random sampling). In a number of states they performed stratified sampling that allowed them to estimate prevalence for regions within their respective states.

I think that targeting only people with landline telephones introduces a source of bias to the sample because they were unable to reach a huge segment of the population without landline telephones. Later in 2008 they introduced cellphone surveys, so they could at least reach people who have cellphones but no landline. I think the latter could introduce duplicates to the data. Also, there is still a small source of bias, since people without both landline and cellphone are excluded from the study.

Since the observational study uses random sampling, results obtained can be generalizable to the United States non-institutionalized adult population, aged 18 years or older. Only correlation/association statements can be generalized to this population. Since there is no random assignment of any treatment to cases, no causal statements can be made about the data.

Below we show a general overview of the variables of interest for our three research questions. We can see the majority of them are categorical variables and just a pair of them are numerical discrete variables.

```{r}
brfss2013 %>%
  select(lsatisfy, X_educag, marital, genhlth, scntwrk1, sex, qlstres2, X_chldcnt) %>%
  str()
```

* * *

## Part 2: Research questions

### Q1: Explore retationship between satisfaction with life, education level and marital status.
As a first question we are interested in exploring the relationship between satisfaction with life (`lsatisfy`) and education level (`X_educag`) with respect to different marital status (`marital`). We select the columns of interest and take a closer look at them. All three variables should be categorical and for this particular case we should discard "NA" observations from the analysis since we are interested in people that actually answered to the questions related to each variable.

```{r count_lsatisfy}
brfss2013 %>% 
  group_by(lsatisfy) %>%
  summarise(count = n())
```

```{r count_after_remove_nas_lsatisfy}
brfss2013 %>% 
  filter(!is.na(lsatisfy)) %>%
  group_by(lsatisfy) %>%
  summarise(count = n())
```

```{r count_X_educag}
brfss2013 %>% 
  group_by(X_educag) %>%
  summarise(count = n())
```

```{r count_after_remove_nas_X_educag}
brfss2013 %>% 
  filter(!is.na(X_educag)) %>%
  group_by(X_educag) %>%
  summarise(count = n())
```

The marital status categorical variable has too many levels. We'll simplify it and keep only two levels. We'll keep the `Married` level as it is, and merge the rest of the 5 levels as `Not Married`. We'll also remove missing values.

```{r count_marital}
brfss2013 %>% 
  group_by(marital) %>% 
  summarise(count = n())
```

```{r count_after_remove_nas_marital}
brfss2013 %>% 
  filter(!is.na(marital)) %>%
  group_by(marital) %>% 
  summarise(count = n())
```

```{r group_marital_levels}
brfss2013 <- brfss2013 %>% 
  mutate(simple_marital = ifelse(marital == "Married", "Married", "Not Married")) %>%
  mutate(simple_marital = as.factor(simple_marital))

brfss2013 %>% 
  group_by(simple_marital) %>%
  summarise(count = n())
```

We'll create a small clean dataset with no missing values, consisting of only the three features we are interested in, related to question 1 of this exploratory analysis. 

```{r create_q1_small_dataset}
small_brfss2013_q1 <- brfss2013 %>% 
  select(lsatisfy, X_educag, simple_marital) %>%
  filter(!is.na(lsatisfy), !is.na(X_educag), !is.na(simple_marital))
```

Let's make a quick overview of the datasets again to check they were correctly constructed.

```{r review_q1_small_dataset}
small_brfss2013_q1 %>%
  group_by(lsatisfy, X_educag, simple_marital) %>%
  summarise(count = n())
```

### Q2: Explore retationship between general health, hours per week worked and sex.
The second question is related to exploring the relationship between general health (`genhlth`) and hours per week worked (`scntwrk1`) with respect to gender (`sex`). We become familiar with the features of interest and take a closer look at them. In this case we'll also be discarding "NA" observations from the analysis.

```{r count_genhlth}
brfss2013 %>% 
  group_by(genhlth) %>%
  summarise(count = n())
```

```{r count_after_remove_nas_genhlth}
brfss2013 %>% 
  filter(!is.na(genhlth)) %>%
  group_by(genhlth) %>%
  summarise(count = n())
```

Feature `scntwrk1` has numeric discrete values, so we would want to take a look at numerical summaries such as means, medians, etc. It's important to first filter "NA" values before calculating these measures.

```{r str_scntwrk1}
str(brfss2013$scntwrk1)
```

```{r numeric_summary_after_remove_nas_scntwrk1}
brfss2013 %>% 
  filter(!(is.na(scntwrk1))) %>%
  summarise(hoursperweekmean = mean(scntwrk1), 
            hoursperweekmedian = median(scntwrk1), 
            hoursperweeksd = sd(scntwrk1), 
            hoursperweekmin = min(scntwrk1), 
            hoursperweekmax = max(scntwrk1))
```

We can see the mean value is reasonable, since the most commmon amount of hours of work per week for a regular worker is 40 hours (8 hours a day, 5 days a week). We think the maximum value is a bit too big, but is still a valid number, because is less than 24 hours a day, regardless of the person working 5 or 7 days a week. We also know this value could vary a lot depending on the type of work the person is doing, so we'll asume there are no obvious errors in the data.

```{r count_sex}
brfss2013 %>% 
  group_by(sex) %>%
  summarise(count = n())
```

```{r count_after_remove_nassex}
brfss2013 %>% 
  filter(!is.na(sex)) %>%
  group_by(sex) %>%
  summarise(count = n())
```

We'll again create a small clean dataset with no missing values, consisting of only the three features we are interested, related to question 2 of this exploratory analysis. 

```{r create_q2_small_dataset}
small_brfss2013_q2 <- brfss2013 %>% 
  select(genhlth, scntwrk1, sex) %>%
  filter(!is.na(genhlth), !is.na(scntwrk1), !is.na(sex))
```

Checking the dataset, we can see it was correctly created.

```{r review_q2_small_dataset}
small_brfss2013_q2 %>%
  group_by(genhlth, sex) %>%
  summarise(count = n())

small_brfss2013_q2 %>%
  select(scntwrk1) %>%
  summarise(hoursperweekmean = mean(scntwrk1), 
            hoursperweekmedian = median(scntwrk1), 
            hoursperweeksd = sd(scntwrk1), 
            hoursperweekmin = min(scntwrk1), 
            hoursperweekmax = max(scntwrk1))
```

### Q3: Explore retationship between anxiety in the past 30 days and number of children in household.
Last but not least, the third question is related to exploring the relationship between the anxiety in the past 30 days (`qlstres2`) with the number of children in the household (`X_chldcnt`). We take a look at the features of interest and discarding "NA" observations from the analysis if needed.

```{r str_qlstres2}
str(brfss2013$qlstres2)
```

```{r count_qlstres2}
brfss2013 %>% 
  group_by(qlstres2) %>%
  summarise(count = n())
```

We can see there are several missing values and that the possible values range from 0 to 10. This variable has stored its values as integers, but since it has a finite number of possible values we'll transform it to a categorical variable with 3 levels. We'll merge responses corresponding to values from 0 to 3 into level `Low stress`, values from 4 to 6 into level `Regular stress` and finally values between 7 and 10 into level `High stress`.

```{r as_factor_qlstres2}
# brfss2013 <- brfss2013 %>%
#   filter(!is.na(qlstres2)) %>%
#   mutate(stress_past_30d = ifelse(qlstres2 < 4, "Low stress", ifelse(qlstres2 < 7, "Regular stress", "High stress"))) %>%
#   mutate(catstress_past_30d = as.factor(stress_past_30d))

brfss2013 <- brfss2013 %>%
  mutate(stress_past_30d = ifelse(qlstres2 < 4, "Low stress", ifelse(qlstres2 < 7, "Regular stress", "High stress"))) %>%
  mutate(stress_past_30d = as.factor(stress_past_30d))
```

```{r count_catqlstres2}
brfss2013 %>% 
  filter(!is.na(stress_past_30d)) %>%
  group_by(stress_past_30d) %>%
  summarise(count = n())
```

```{r count_X_chldcnt}
brfss2013 %>% 
  group_by(X_chldcnt) %>%
  summarise(count = n())
```

```{r count_after_remove_nasX_chldcnt}
brfss2013 %>% 
  filter(!is.na(X_chldcnt)) %>%
  group_by(X_chldcnt) %>%
  summarise(count = n())
```

Same as we've done for the other two research questions, We'll create a small clean dataset with no missing values, consisting of only the three features we are interested in, related to question 3 of this exploratory analysis. 

```{r create_q3_small_dataset}
small_brfss2013_q3 <- brfss2013 %>% 
  select(stress_past_30d, X_chldcnt) %>%
  filter(!is.na(stress_past_30d), !is.na(X_chldcnt))
```

```{r review_q3_small_dataset}
small_brfss2013_q3 %>%
  group_by(stress_past_30d, X_chldcnt) %>%
  summarise(count = n())
```

We can see that after in the process of creating the third small dataset, there are no samples belonging to the level `Five or more children in household` of categorical variable `X_chldcnt`. We can also see that the last two levels of the same variable (`Three children in household` and `Four children in household`) have very few observations, so we'll proceed to merge them in a single level we'll name `More than two children in household`.

```{r merge_levels_X_chldcnt_q3_dataset}
small_brfss2013_q3$X_chldcnt <- small_brfss2013_q3$X_chldcnt %>% fct_collapse("More than two children in household" = c("Three children in household", "Four children in household"))

small_brfss2013_q3$X_chldcnt <- factor(small_brfss2013_q3$X_chldcnt)

small_brfss2013_q3 %>%
  group_by(stress_past_30d, X_chldcnt) %>%
  summarise(count = n())
```

* * *

## Part 3: Exploratory data analysis

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button (green button with orange arrow) above. Make sure that your code is visible
in the project you submit. Delete this note when before you submit your work.

**Research quesion 1:**

```{r q1_graph1}
ggplot(small_brfss2013_q1, aes(x = lsatisfy, fill = X_educag)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Satisfaction with life") +
  labs(fill = "Education level") +
  geom_bar(position = "fill")
```

```{r q1_contingency_table}
q1table <- table(small_brfss2013_q1$lsatisfy, small_brfss2013_q1$X_educag)
margin.table(q1table, 1) # row sums (lsatisfy)
margin.table(q1table, 2) # column sums (X_educag)

prop.table(q1table)
prop.table(q1table, 1) # row sums (lsatisfy)
prop.table(q1table, 2) # column sums (X_educag)
```

```{r q1_graph2}
ggplot(small_brfss2013_q1, aes(x = lsatisfy, fill = X_educag)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Satisfaction with life") +
  labs(fill = "Education level") +
  geom_bar(position = "fill") + 
  facet_wrap( ~ simple_marital, ncol = 2)
```


# ```{r algo}
# ```

<!-- Since the proportion of spam changes across the groups in Figure 1.39(b), we can conclude the variables are dependent, which is something we were also able to discern using table proportions. -->

<!-- I think the idea is that your hypothesis (relationship) appears to be true, but in reality there is a confounding variable which might be the real explanation. For example: -->

<!-- Hypothesis: older people are less happy. -->

<!-- EDA to undertake: describe relationship between age and happiness rating. -->

<!-- Posslbe conclusion: older people do seem to be less happy. -->

<!-- Confounding (third) variable: older people are more likely to be female than male. -->

<!-- Further investigation: are females less happy than males? If yes, can this explain the previous conclusion? -->

**Research quesion 2:**

```{r}
small_brfss2013_q2 %>% group_by(genhlth, sex) %>%
  summarise(hpw_worked_median = mean(scntwrk1)) %>%
  ggplot(aes(x = genhlth, y = hpw_worked_median, fill = genhlth)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(x = "General health") +
    labs(fill = "General health") + 
    labs(y = "Hours per week worked") +
    geom_bar(stat = "identity") + 
    facet_wrap( ~ sex, ncol = 2)
```



**Research quesion 3:**

```{r}
small_brfss2013_q3 %>%
  group_by(stress_past_30d, X_chldcnt) %>%
  summarise(count = n())
```

```{r}
ggplot(small_brfss2013_q3, aes(x = X_chldcnt, fill = stress_past_30d)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Num. of children in household") +
  labs(fill = "Level of anxiety in past 30 days") + 
  geom_bar(position = "fill")
```

